---
title: "Study of different bin numbers"
author: "Quentin D. Read"
date: "September 21, 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Summary

The issue we have is that we want to visualize the distribution of energy by rodent mass class on a log-log plot. This requires log-binning to adjust for the different relative widths of the mass bins in that space. We want to be able to compare the different NEON sites to one another so it would be ideal if all sites have the same bin boundaries. However the range of masses is very different from site to site. Therefore if all bin boundaries are set to be the same for each site, we end up with some sites having very few bins with a nonzero value because their range is so restricted. If the number of bins is set too high, there is a different problem: there start to be gaps in between nonzero bins. Here are some alternatives for how to deal with this problem.

# Range of masses at each site

Load the pre-processed mammal data and take a look at the range of values we get at each site. Consider only 2016 sampling of sites with at least 50 individuals and 4 species (what we are using for other analyses).

```{r}
library(ggplot2)
library(dplyr)
library(lubridate)

source('~/Documents/GitHub/NEON_repos/rodentee/load_mammal_data_withimputed.r')

m16 <- mammal_data %>%
  filter(year == 2016) %>%
  group_by(siteID) %>%
  filter(n() >= 50, length(unique(taxonID)) >= 4) 

p_range <- m16 %>%
  summarize(min_mass = min(weight), max_mass = max(weight)) %>%
  ggplot(aes(x = siteID, xend = siteID, y = min_mass, yend = max_mass)) +
  geom_segment() +
  labs(y = 'mass (g)', x = 'site') +
  coord_flip() +
  theme_bw()

p_range + scale_y_log10()

```

# Show bin boundaries

Here are some different numbers of bin boundaries. We set the bin boundaries to be evenly spaced on a logarithmic scale between the minimum and maximum mass value observed across all sites (5 to 436 grams).

```{r}
(mass_range <- range(m16$weight))
```

```{r}
# Function to find number of individuals in each bin
# (not necessary to actually do the log binning here)
bin_table <- function(x, b) {
  ints <- findInterval(x, b, rightmost.closed = TRUE)
  count <- sapply(1:n_bins, function(i) sum(ints==i))
  midpoint <- 10^(log10(b)[-1] - diff(log10(b))/2)
  data.frame(midpoint, count)
}

# Function to make a plot of the numbers of individuals
bin_count_plot <- function(bins) {
  ggplot(bins, aes(x = siteID, y = midpoint, label = count)) +
    geom_text(aes(color = count > 0)) +
    scale_y_log10() +  
    labs(y = 'mass (g)', x = 'site') +
    scale_color_manual(values = c('gray50','red')) +
    geom_hline(data = data.frame(b = bin_bounds), 
               aes(yintercept = b),
               color = 'gray75',
               linetype = 'dotted') +
    theme_bw() +
    theme(panel.grid = element_blank(),
          legend.position = 'none') +
    coord_flip()
}
```

What if we binned every site into 10 bins on that axis? We will need 11 bin boundary numbers.

```{r}
n_bins <- 10

(bin_bounds <- exp(seq(log(mass_range[1]), log(mass_range[2]), length.out = n_bins + 1)))
```

Look at how many individuals appear in each interval at each site with that number of bins. If there are zero individuals, there will be no value for the bin at that site.


```{r}
bin10 <- m16 %>%
  do(bin_table(.$weight, bin_bounds))

bin_count_plot(bin10)
```

As you can see if we divide evenly by 10 bins in log space, we rarely get all 10 to be nonzero at any given site. Many sites have only 3 or 4 bins because they actually have a very restricted range of masses.

Let's try it with 20 bins now.

```{r}
n_bins <- 20
bin_bounds <- exp(seq(log(mass_range[1]), log(mass_range[2]), length.out = n_bins + 1))

bin20 <- m16 %>%
  do(bin_table(.$weight, bin_bounds))

bin_count_plot(bin20)
```

We start to get gaps, and the sites with less range still only have around 5 or 6 bins. Not ideal.

# New binning algorithm

We want as many points to line up across sites as possible so the new algorithm works as follows: bin the masses given the original input bin boundaries and check how many bins have nonzero values. As long as the number of nonzero values is less than the desired number of bins $n$, we split the bin with the most individuals in half, adjusting for the now reduced width of the two bins created from one. Rebin and repeat until $n$ bins have at least one individual in them.

```{r}
logbin_split <- function(x, y = NULL, bin_edges) {
  logx <- log10(x)                                           # log transform x value (biomass)
  bin_edges <- log10(bin_edges)
  
  logxbin <- findInterval(logx, bin_edges, rightmost.closed = TRUE) # assign each individual to a bin
  
  n <- length(bin_edges) - 1 # Desired number of nonzero bins
  
  # While loop to keep splitting the biggest bin in half until we get n nonzero bins
  while(length(unique(logxbin)) < n) {
    uniquebin <- unique(logxbin)
    n_mode <- uniquebin[which.max(tabulate(match(logxbin, uniquebin)))] # Which logbin is the mode (most individuals)?
    # Add a midpoint in the middle of that bin.
    bin_edges <- c(bin_edges[1:n_mode], mean(bin_edges[n_mode:(n_mode+1)]), bin_edges[(n_mode+1):length(bin_edges)])
    logxbin <- findInterval(logx, bin_edges, rightmost.closed = TRUE)
  }
  
  n <- length(bin_edges) - 1 # New n
  
  bin_midpoints <- 10^(bin_edges[-1] - diff(bin_edges)/2)
  bin_widths <- diff(10^bin_edges)                              # get linear width of each bin
  bin_factor <- factor(logxbin, levels=1:n)                  # convert bin to factor (required to deal with zeroes if present)
  bin_counts <- table(bin_factor)                            # find number of individuals in each bin
  if (!is.null(y)) {
    rawy <- tapply(y, bin_factor, sum)                       # sum y value (production) in each bin
    rawy[is.na(rawy)] <- 0                                   # add zeroes back in if present
    bin_values <- as.numeric(rawy/bin_widths)                # divide production by width for each bin 
  }
  else {
    bin_values <- as.numeric(bin_counts/bin_widths)          # 1-dimensional case.
  }
  
  return(data.frame(bin_midpoint = bin_midpoints,            # return result!
                    bin_value = bin_values,                  # also add bin min and max for bar plot purposes
                    bin_count = as.numeric(bin_counts),
                    bin_min = 10^bin_edges[1:n],
                    bin_max = 10^bin_edges[2:(n+1)]))
}

```

This is the old algorithm for comparison.

```{r}
logbin_setedges <- function(x, y = NULL, bin_edges) {
  logx <- log10(x)                                           # log transform x value (biomass)
  bin_edges <- log10(bin_edges)
  
  logxbin <- findInterval(logx, bin_edges, rightmost.closed = TRUE) # assign each individual to a bin
  
  n <- length(bin_edges) - 1
  
  bin_midpoints <- 10^(bin_edges[-1] - diff(bin_edges)/2)
  bin_widths <- diff(10^bin_edges)                              # get linear width of each bin
  bin_factor <- factor(logxbin, levels=1:n)                  # convert bin to factor (required to deal with zeroes if present)
  bin_counts <- table(bin_factor)                            # find number of individuals in each bin
  if (!is.null(y)) {
    rawy <- tapply(y, bin_factor, sum)                       # sum y value (production) in each bin
    rawy[is.na(rawy)] <- 0                                   # add zeroes back in if present
    bin_values <- as.numeric(rawy/bin_widths)                # divide production by width for each bin 
  }
  else {
    bin_values <- as.numeric(bin_counts/bin_widths)          # 1-dimensional case.
  }
  
  return(data.frame(bin_midpoint = bin_midpoints,            # return result!
                    bin_value = bin_values,                  # also add bin min and max for bar plot purposes
                    bin_count = as.numeric(bin_counts),
                    bin_min = 10^bin_edges[1:n],
                    bin_max = 10^bin_edges[2:(n+1)]))
  
}

```


Here the algorithm is applied to the 2016 mammal data, ensuring all sites have 10 nonzero bins.

```{r}
n_bins <- 10

bin_bounds <- exp(seq(log(mass_range[1]), log(mass_range[2]), length.out = n_bins + 1))

m16split <- m16 %>%
  do(logbin_split(.$weight, bin_edges = bin_bounds))

ggplot(m16split, aes(x = siteID, y = bin_midpoint, label = bin_count)) +
    geom_text(aes(color = bin_count > 0), size = 3) +
    scale_y_log10() +  
    labs(y = 'mass (g)', x = 'site') +
    scale_color_manual(values = c('gray50','red')) +
    theme_bw() +
    theme(panel.grid = element_blank(),
          legend.position = 'none') +
    coord_flip()
```

Now as you can see, there are as many bins lined up with each other as possible, and there are guaranteed to be 10 values for each site.

# Show the old and new methods side by side

```{r}
m16oldbin <- m16 %>%
  do(logbin_setedges(.$weight, bin_edges = bin_bounds))

allbin <- rbind(data.frame(method = 'old', m16oldbin),
                data.frame(method = 'new', m16split))

ggplot(allbin %>% filter(bin_value > 0), 
       aes(x = bin_midpoint, y = bin_value, color = method, shape = method)) +
  geom_point(size = 2, alpha = 0.5) +
  facet_wrap(~ siteID) +
  scale_x_log10(name = 'mass (g)') +
  scale_y_log10(name = 'density (1/g)') +
  theme_bw() +
    theme(panel.grid = element_blank(),
          legend.position = 'bottom')
```

