---
title: "NEON rodent energy equivalence"
author: "Quentin D. Read"
date: "May 11, 2018"
output: pdf_document
---

# Define function for binning

Ethan White's log binning algorithm, same as we used previously.

```{r}
logbin_setedges <- function(x, y = NULL, bin_edges) {
  logx <- log10(x)                                           # log transform x value (biomass)
  bin_edges <- log10(bin_edges)
  n <- length(bin_edges) - 1
  logxbin <- rep(NA, length(logx))                           # create data structure to assign trees to bins
  b <- bin_edges                                             # add a little to the biggest bin temporarily
  b[length(b)] <- b[length(b)] + 1                           # (so that the biggest single tree is put in a bin)
  for (i in 1:length(logx)) {
    logxbin[i] <- sum(logx[i] >= b)                          # assign each tree to a bin
  }
  bin_midpoints <- 10^(bin_edges[-1] - diff(bin_edges)/2)
  bin_widths <- diff(10^bin_edges)                              # get linear width of each bin
  bin_factor <- factor(logxbin, levels=1:n)                  # convert bin to factor (required to deal with zeroes if present)
  bin_counts <- table(bin_factor)                            # find number of trees in each bin
  if (!is.null(y)) {
    rawy <- tapply(y, bin_factor, sum)                       # sum y value (production) in each bin
    rawy[is.na(rawy)] <- 0                                   # add zeroes back in if present
    bin_values <- as.numeric(rawy/bin_widths)                # divide production by width for each bin 
  }
  else {
    bin_values <- as.numeric(bin_counts/bin_widths)          # 1-dimensional case.
  }
  
  return(data.frame(bin_midpoint = bin_midpoints,            # return result!
                    bin_value = bin_values,                  # also add bin min and max for bar plot purposes
                    bin_count = as.numeric(bin_counts),
                    bin_min = 10^bin_edges[1:n],
                    bin_max = 10^bin_edges[2:(n+1)]))
  
}

```

# Load data

This CSV was pulled from NEON in April 2018. I deleted some unneeded columns and rows denoting empty traps so that the file is smaller but other than that it isn't modified at all from what I got directly from NEON.

```{r, message=FALSE}
# Loaded from already-pulled CSV.
mammal_data <- read.csv('C:/Users/Q/google_drive/NEON_EAGER/Manuscript5_RodentEE/mammal_reduced.csv', stringsAsFactors = FALSE)
```

# Data processing

Get rid of all recaptures. *Edit: originally I removed juveniles but they are included again. Only ~10% are juveniles.* Other than that do no other quality control for now. Probably a good idea to look back through later and get rid of odd outliers. I chose to do 2016 for this analysis because it has the most data (about 10000 individuals). Get rid of any row without weight (leaving ~8000 rows). Also keep only the columns we care about for the moment. 

```{r, message = FALSE}
library(dplyr)
library(lubridate)

mammal_data <- mammal_data %>%
  filter(!recapture %in% c('U','Y')) %>%
  #filter(!lifeStage %in% c('juvenile')) %>%
  mutate(month = month(collectDate), year = year(collectDate)) %>%
  filter(year == 2016, !is.na(weight)) %>%
  select(domainID, siteID, plotID, collectDate, month, year, taxonID, lifeStage, weight)
  

```

Create a rough guess for relative metabolic rate (energy flux) by taking $mass^{\frac{3}{4}}$ for each individual.

```{r}
mammal_data <- mammal_data %>%
  mutate(relativeEnergy = weight ^ 0.75)
```

# Binning

Pool all the measurements together to get the absolute max and min to be used for binning each site. 

```{r}
range(mammal_data$weight)
# One is zero so let's get rid of it.
mammal_data <- filter(mammal_data, weight > 0)
(weight_range <- range(mammal_data$weight))

n_bins <- 8

# If we did 8 bins these would be the bin bounds.
(bin_bounds <- exp(seq(log(weight_range[1]), log(weight_range[2]), length.out = n_bins + 1)))
```

Use the log-binning algorithm to bin each site's body mass measurements based on the bin bounds we just determined. We don't care about the date of sampling or the plot within site that's sampled. They are all lumped.

```{r}
mass_bins <- mammal_data %>%
  group_by(siteID) %>%
  do(logbin_setedges(x = .$weight, bin_edges = bin_bounds))
```

Use the same bins to get the binned energy flux. 

```{r}
energy_bins <- mammal_data %>%
  group_by(siteID) %>%
  do(logbin_setedges(x = .$weight, y = .$relativeEnergy, bin_edges = bin_bounds))
```

Make a plot, by site, of binned body mass. I only plotted the ones with at least 100 individuals for now.

```{r, message = FALSE}
library(ggplot2)

n_individuals <- mass_bins %>%
  group_by(siteID) %>%
  summarize(n = sum(bin_count)) %>%
  filter(n >= 100)

mass_bins %>% 
  filter(bin_value > 0, siteID %in% n_individuals$siteID) %>%
ggplot(aes(x = bin_midpoint, y = bin_value)) + 
  facet_wrap(~ siteID) +
  theme_bw() + geom_line() + geom_point() +
  scale_x_log10() + scale_y_log10()
```

Make a plot, by site, of binned (relative) total energy flux (assuming each individual has a metabolic rate exactly proportional to 3/4 power of its mass). Again only sites with at least 100 individuals are shown. The patterns look fairly messy.

```{r}
energy_bins %>% 
  filter(bin_value > 0, siteID %in% n_individuals$siteID) %>%
ggplot(aes(x = bin_midpoint, y = bin_value)) + 
  facet_wrap(~ siteID) +
  theme_bw() + geom_line() + geom_point() +
  scale_x_log10(name = 'Mass (g)') + scale_y_log10(name = 'Total relative energy')
```

Boxplot showing distributions of nonzero bin values across sites.

```{r}
energy_bins %>% 
  filter(siteID %in% n_individuals$siteID) %>%
  ggplot(aes(x = bin_midpoint, y = bin_value)) +
  stat_boxplot(aes(group = bin_midpoint)) +
  scale_x_log10(name = 'Mass (g)') + scale_y_log10(name = 'Total relative energy') +
  theme_bw()
```

# Binning by species richness

I modified the log binning function to get the number of species in each bin rather than the total energy. This does not use the mean value by species but the mass value for each individual. Therefore a species could be counted in multiple bins.

```{r}
logbin_categorical <- function(x, y, bin_edges) {
  xbin <- cut(x, breaks = bin_edges, include.lowest = TRUE)
  data.frame(x, y, xbin) %>%
    group_by(xbin) %>%
    summarize(richness = length(unique(y)))
}
```

Here the binning is done.

```{r}
bin_mids <- exp(log(bin_bounds)[-1] - diff(log(bin_bounds))/2)

mammal_richness_bins <- mammal_data %>%
  group_by(siteID) %>%
  do(logbin_categorical(x = .$weight, y = .$taxonID, bin_edges = bin_bounds))
```

Plots separated by site (species richness on the y axis is not on a log scale).

```{r, warning = FALSE}
mammal_richness_bins %>% 
  mutate(bin_midpoint = bin_mids[as.numeric(xbin)]) %>%
  filter(siteID %in% n_individuals$siteID) %>%
  ggplot(aes(x = bin_midpoint, y = richness)) + 
  facet_wrap(~ siteID) +
  theme_bw() + geom_line() + geom_point() +
  scale_x_log10(name = 'Mass (g)') + scale_y_continuous(name = 'Species richness')
```

Boxplot showing richness distribution for each bin across sites.

```{r, warning = FALSE}
mammal_richness_bins %>% 
  mutate(bin_midpoint = bin_mids[as.numeric(xbin)]) %>%
  filter(siteID %in% n_individuals$siteID) %>%
  ggplot(aes(x = bin_midpoint, y = richness)) +
  stat_boxplot(aes(group = bin_midpoint)) +
  scale_x_log10(name = 'Mass (g)') + scale_y_continuous(name = 'Species richness') +
  theme_bw()
```